Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-06 workflow/workflow.Rmd
Updates to text.

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley; Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA

@@

This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017]. The olfactory epithelium contains mature olfactory sensory neurons (mOSN) that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium. When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs), which become activated to differentiate and reconstitute all major cell types in the epithelium.

@@

We need the following packages.

@@

Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor R package `zinbwave`. The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a sample-level intercept, which serves as a global-scaling normalization factor. The user can also include both gene-level and sample-level covariates. The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects). A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema). For greater detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

@@

As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects.

@@

Note that the defaults in `RSEC` are designed for input data that are the actual (normalized) counts. Here, we are applying `RSEC` instead to the low-dimensional `W` matrix of the previous step, for which we make a separate `SummarizedExperiment` object. For this reason, we choose to not use certain options in `RSEC`. In particular, we do not use the default dimensionality reduction step, since our input `W` is already in a space of reduced dimension. Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`dimReduce`) and the construction of the hiearchy between the clusters (`dendroReduce`). We also skip the option to merge our clusters based on the amount of differential gene expression between clusters.

@@

```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering, color-coded by RSEC clustering."}
plotBarplot(ceObj, legend = F) 
```

@@

[FP: Elizabeth or Davide, how can we change the colors in Figure \@ref(fig:addPublishedClusters) to match the colors we use for publishedClusters (that is `col_clus` like in Figure \@ref(fig:mdsW)). I tried colPalette argument, but it does not work with SummarizedExperiment input. Also, would it be possible to have the same order and labels for the barplots in Figures \@ref(fig:barplotOurs) and \@ref(fig:addPublishedClusters) (barplots labelled -1 and Not Assigned are the same, right?)?

```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering, color-coded by original published clustering."}
ceObj <- addClusters(ceObj, colData(ceObj)$publishedClusters, 
                     clusterLabel = "publishedClusters")
plotBarplot(ceObj, whichClusters=c("combineMany","publishedClusters"),
            xlab = "", legend = F) 
```

@@

```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clusters."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
     xlab = "Component1", ylab = "Component2")
legend(x = "bottomright", legend = names(pal), cex = .5,
       fill = pal, title = "Sample")
```

@@

The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`, respectively. The first takes a low-dimensional representation of the cells and a vector of cluster labels. It fits an MST to the clusters and identifies lineages as paths through this tree. The output of `getLineages` is an object of class `SlingshotDataSet` containing all the information used to fit the tree and identify lineages. The function `getCurves` then takes this object as input and fits simultaneous principal curves to the identified lineages. These functions can be run separately, as below, or jointly by the wrapper function `slingshot`.

@@

Cells in cluster `c4` have a cluster label of `-2` in the original published clustering meaning that they were not assigned to any cluster in the original publication. These cells were actually identified as non-sensory contaminants as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)) and were removed from the original published clustering. While it is reassuring that our workflow clustered these cells separately with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes as cells in this cluster do not participate in the cell differentiation process. Note that our workflow's clustering did not capture all the cells overexpressing gene `Reg3g` in cluster `c4` (see Figure \@ref(fig:boxplotReg3g)), however we keep these cells to infer lineages as they did not seem to influence the clustering.

[FP: check writing + table with correspondance of clusters]

```{r boxplotReg3g, fig.width=4, fig.height=4, fig.cap="Boxplots of the log counts of gene Reg3g for cells in cluster c4 (left boxplot) and cells not in cluster c4 (right boxplot)."}
c4 <- rep("other clusters", ncol(se)) 
c4[primaryClusterNamed(ceObj) == "c4"] <- "cluster c4"
boxplot(log1p(assay(se)["Reg3g", ]) ~ c4, ylab = "log counts", cex.axis = .8, cex.lab = .8)
```

@@

```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred global lineage structure."}
pairs(lineages, type="lineages", col = pal[cl])
```

@@

```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)
pairs(lineages, type="curves", col = pal[cl])
```

@@

In an effort to improve scRNA-seq data analysis workflows, we are currently exploring a variety of applications and extensions of our ZINB-WaVE model. In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference. In addition, we are extending ZINB-WaVE to differentially expression, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns. We are also developing a method to identify genes that are DE either within or between lineages inferred from Slingshot.

>>>>

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA

@@

This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017]. The olfactory epithelium contains mature olfactory sensory neurons (mOSN) that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBC), which are the actively proliferating cells in the epithelium. When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBC), which become activated to differentiate and reconstitute all major cell types in the epithelium.

@@

The following packages are needed.

@@

Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor R package `zinbwave`. The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a cell-level intercept, which serves as a global-scaling normalization factor. The user can also specify both gene-level and cell-level covariates. The inclusion of observed and unobserved cell-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects). A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema). For greater detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

@@

As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`.

@@

Note that the defaults in `RSEC` are designed for input data that are the actual (normalized) counts. Here, we are applying `RSEC` instead to the low-dimensional `W` matrix from ZINB-WaVE, for which we make a separate `SummarizedExperiment` object. For this reason, we choose to not use certain options in `RSEC`. In particular, we do not use the default dimensionality reduction step, since our input `W` is already in a space of reduced dimension. Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`dimReduce`) and the construction of the hiearchy between the clusters (`dendroReduce`). We also skip the option to merge our clusters based on the amount of differential gene expression between clusters.

@@

```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster for our workflow's RSEC clustering."}
plotBarplot(ceObj, legend = F) 
```

@@

[FP: Elizabeth or Davide, how can we change the colors in Figure \@ref(fig:addPublishedClusters) to match the colors we use for publishedClusters (that is `col_clus` like in Figure \@ref(fig:mdsW)). I tried colPalette argument, but it does not work with SummarizedExperiment input. Also, would it be possible to have the same order and labels for the barplots in Figures \@ref(fig:barplotOurs) and \@ref(fig:addPublishedClusters) (barplots labelled -1 and Not Assigned are the same, right?)?]

```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's RSEC clustering, color-coded by original published clustering."}
ceObj <- addClusters(ceObj, colData(ceObj)$publishedClusters, 
                     clusterLabel = "publishedClusters")
plotBarplot(ceObj, whichClusters=c("combineMany","publishedClusters"),
            xlab = "", legend = F) 
```

@@

```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
     xlab = "Component1", ylab = "Component2")
legend(x = "bottomright", legend = names(pal), cex = .5,
       fill = pal, title = "Sample")
```

@@

The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`, respectively. The first takes as input a low-dimensional representation of the cells and a vector of cluster labels. It fits an MST to the clusters and identifies lineages as paths through this tree. The output of `getLineages` is an object of class `SlingshotDataSet` containing all the information used to fit the tree and identify lineages. The function `getCurves` then takes this object as input and fits simultaneous principal curves to the identified lineages. These functions can be run separately, as below, or jointly by the wrapper function `slingshot`.

@@

Cells in cluster `c4` have a cluster label of `-2` in the original published clustering, meaning that they were not assigned to any cluster. These cells were actually identified as non-sensory contaminants, as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)), and were removed from the original published clustering. While it is reassuring that our workflow clustered these cells separately, with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes, as cells in this cluster do not participate in the cell differentiation process. Note that our workflow's clustering only captured XXX out of XXX cells overexpressing gene `Reg3g` in cluster `c4` (see Figure \@ref(fig:boxplotReg3g)), however we retain the remaining cells to infer lineages as they did not seem to influence the clustering.

[FP: check writing + table with correspondance of clusters. SD: I did minor edits to text above. Please check. Perhaps mention total number of contaminant cells and how many are in cluster `c4`.]

```{r boxplotReg3g, fig.width=4, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g for cells in cluster c4 (left) and cells not in cluster c4 (right)."}
c4 <- rep("other clusters", ncol(se)) 
c4[primaryClusterNamed(ceObj) == "c4"] <- "cluster c4"
boxplot(log1p(assay(se)["Reg3g", ]) ~ c4, ylab = "log counts", cex.axis = .8, cex.lab = .8)
```

@@

```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
pairs(lineages, type="lineages", col = pal[cl])
```

@@

```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)
pairs(lineages, type="curves", col = pal[cl])
```

@@

In an effort to improve scRNA-seq data analysis workflows, we are currently exploring a variety of applications and extensions of our ZINB-WaVE model. In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference. In addition, we are extending ZINB-WaVE to identify differentially expressed genes, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns. We are also developing a method to identify genes that are DE either within or between lineages inferred from Slingshot.
